% Vulnerability Macro
\vuln
    { %title
      Documentation enhancements
    }
    { % CVSS RATING
      0.0
    }
    { % CVSS Calculator Link
    https://nvd.nist.gov/vuln-metrics/cvss/v3-calculator?vector=AV:N/AC:L/PR:N/UI:N/S:C/C:N/I:N/A:N/E:U/RL:X/RC:X/CR:H/IR:H/AR:M/MAV:X/MAC:X/MPR:X/MUI:X/MS:X/MC:N/MI:N/MA:N&version=3.1
    }
    {
    None
    }
    { % vulnerability types
    \item insufficient-documentation
    \item incorrect-documentation
    }
    { % Description 
    Proper documentation is crucial to avoid introducing new bugs in the future.
Codebase generally contains few comments, but the consistent namings and module structure together with self-commenting validator checks make it a lesser problem.
Following are issues that could be improved still:
\begin{itemize}
    \item \texttt{RecordEpochToScaleToSum} endpoint is missing from the yellowpaper
    \item Minimal collateral ratio is passed in code as a percentage. This is undocumented both in the types and functions using it.
    \item \texttt{sessSnapshot} field is confusingly named. Elsewhere "snapshot" refers to values with \texttt{StabilityPoolSnapshot} type.
    \item Implemention of the tests is poorly documented.
\end{itemize}
    }
    {
    N/A
    }
    { % Recommended Fix
      Fix listed issues.
    }
    { % Links
    \item\href{https://www.atlassian.com/trust/security/security-severity-levels}{Severity Levels for Security Issues}
    \item\href{https://github.com/IndigoProtocol/smart-contracts/issues/478}{Audit - Documentation enhancements. #478}
    }
